---
title: "Value Validation Project: Population Health Analytics"
author: "Walker Blackston, MSPH"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(sqldf)
library(corrplot)
library(gtsummary)
library(gt)
library(caret)
library(pROC)
library(broom)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Executive Summary

> To demonstrate my fit for this team at Ochsner, rather than a traditional cover letter, I am submitting a simulated case study of my abilities. This project simulates a value-based population health strategy leveraging predictive modeling to segment high-risk patients. By integrating clinical and utilization data, we estimate that targeted care management for the top decile could yield a 6.7% ROI within 12 months. This framework is to demonstrate my workflow and approaches to opportunities I might face on this team, rather than specific findings.

# Objectives

Reduce avoidable utilization, target care management, quantify ROI, provide a pilot to transfer to real world data.

# Data & Methods

-   Simulated dataset - we simulated a variety of patient characteristics and claims data in a sample of patients at Ochsner hospitals in the greater New Orleans metro area (n=5,000)
-   A full script of all data generation, output and analysis can be found in an attached Appendix if requested

```{r simulate-data, echo=FALSE, message=FALSE, warning=FALSE}

set.seed(123)

n <- 5000

patient_data <- data.frame(
  patient_id = sprintf("P%04d", 1:n),
  age = round(runif(n, 18, 90)),
  sex = sample(c("Male", "Female"), n, replace = TRUE),
  zip_income = round(rnorm(n, mean = 55000, sd = 16000)),
  hypertension = rbinom(n, 1, 0.4),
  diabetes = rbinom(n, 1, 0.25),
  copd = rbinom(n, 1, 0.15),
  depression = rbinom(n, 1, 0.2),
  cancer = rbinom(n, 1, 0.05),
  ed_visits = rpois(n, lambda = 1.5),
  inpatient_admits = rpois(n, lambda = 0.5),
  primary_care_visits = rpois(n, lambda = 3),
  med_adherence_score = pmin(pmax(rnorm(n, mean = 0.75, sd = 0.15), 0), 1)
)

# Simulated total cost influenced by conditions and utilization
patient_data$total_cost <- round(
  750 +
    500 * patient_data$hypertension +
    700 * patient_data$diabetes +
    900 * patient_data$copd +
    600 * patient_data$depression +
    1500 * patient_data$cancer +
    300 * patient_data$ed_visits +
    1500 * patient_data$inpatient_admits +
    rnorm(n, mean = 1000, sd = 800)
)

risk_score <- with(patient_data,
                   0.02 * age +
                   0.5 * hypertension +
                   0.7 * diabetes +
                   0.8 * copd +
                   0.6 * depression +
                   1.2 * cancer +
                   0.4 * ed_visits +
                   1.0 * inpatient_admits +
                   rnorm(n, 0, 1.5))  # Add noise

# Convert to probability with logistic function
risk_prob <- 1 / (1 + exp(-risk_score))

# Assign high_risk based on probabilistic threshold
patient_data$high_risk <- rbinom(n, 1, prob = risk_prob)

# Preview
head(patient_data)
```

------------------------------------------------------------------------

### Structured Data Management Example (SQL Integration)

While most modeling work is done in R, SQL remains essential for summarizing, cleaning, and staging patient-level data from claims and EMRs. Below is a representative query using `sqldf` to analyze the distribution of high-risk patients by sex and income bracket—showcasing how SDOH proxies might inform any of our stratification strategies.

```{r sql-data-summary, echo=TRUE, message=FALSE}

# Create a derived income bracket for reporting
patient_data$income_bracket <- cut(
  patient_data$zip_income,
  breaks = c(-Inf, 40000, 60000, 80000, Inf),
  labels = c("<40K", "40–60K", "60–80K", "80K+")
)

# SQL-style summary: high-risk patients by sex and income bracket
sql_summary <- sqldf("
  SELECT 
    sex,
    income_bracket,
    COUNT(*) AS total_patients,
    SUM(high_risk) AS high_risk_patients,
    ROUND(100.0 * SUM(high_risk) / COUNT(*), 1) AS pct_high_risk
  FROM patient_data
  GROUP BY sex, income_bracket
  ORDER BY income_bracket, sex
")

sql_summary
```

```{r echo=FALSE}
patient_data %>%
  mutate(high_risk = factor(high_risk, labels = c("Not High Risk", "High Risk"))) %>%
  select(high_risk, age, zip_income, ed_visits, inpatient_admits, med_adherence_score, 
         total_cost, hypertension, diabetes, copd, depression, cancer) %>%
  tbl_summary(by = high_risk,
              statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
              digits = all_continuous() ~ 1,
              label = list(zip_income ~ "ZIP-Level Income",
                           ed_visits ~ "ED Visits",
                           inpatient_admits ~ "Inpatient Admits",
                           med_adherence_score ~ "Medication Adherence Score",
                           total_cost ~ "Total Cost")) %>%
  modify_header(label ~ "**Variable**") %>%
  bold_labels() %>%
  add_p() %>%
  modify_caption("**Table: Patient Characteristics Stratified by Risk Status**")
```

# Exploratory Data Analysis

To better understand the drivers of risk and opportunity for intervention, we conducted exploratory data analysis on the synthetic population.

```{r summary-stats, echo=FALSE}
# Basic summary using gtsummary
patient_data %>%
  select(age, zip_income, ed_visits, inpatient_admits, primary_care_visits, 
         med_adherence_score, total_cost, hypertension, diabetes, copd, depression, cancer) %>%
  tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
              digits = all_continuous() ~ 1,
              label = list(zip_income ~ "ZIP-Level Income",
                           ed_visits ~ "ED Visits",
                           inpatient_admits ~ "Inpatient Admits",
                           primary_care_visits ~ "Primary Care Visits",
                           med_adherence_score ~ "Medication Adherence Score",
                           total_cost ~ "Total Cost",
                           hypertension ~ "HTN",
                           copd ~ "COPD")) %>%
  modify_header(label ~ "**Variable**") %>%
  bold_labels()
```

```{r corr-plot, echo=FALSE}
num_vars <- patient_data %>%
  select(age, zip_income, ed_visits, inpatient_admits, primary_care_visits,
         med_adherence_score, total_cost, hypertension, diabetes, copd, depression, cancer)

corr_matrix <- cor(num_vars)
corrplot(corr_matrix, method = "color", type = "lower", tl.cex = 0.8)
```

-   This **correlation matrix** suggests that total cost is strongly associated with increased inpatient visits as well as moderately associated with the presence of chronic conditions

```{r plots-risk, echo=FALSE}
ggplot(patient_data, aes(x = total_cost, fill = as.factor(high_risk))) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("#002d72", "#fdb813")) + 
  labs(title = "Distribution of Total Cost by Risk Group", x = "Total Cost", fill = "High Risk")

ggplot(patient_data, aes(x = inpatient_admits, fill = as.factor(high_risk))) +
  geom_bar(position = "dodge") + 
  scale_fill_manual(values = c("#002d72", "#fdb813")) + 
  labs(title = "Inpatient Admissions by Risk Group", x = "Inpatient Admissions", fill = "High Risk")
```

-   Visualizations show that a small subset of patients (top 10%) drive a disproportionate share of inpatient admits and total costs
-   High risk groups appear to experience generally higher cost and inpatient admissions than low risk groups

```{r risk-prof-tab, echo=FALSE}
# Create the group-level summary
risk_profile <- patient_data %>%
  group_by(high_risk) %>%
  summarise(
    `Mean Age` = round(mean(age), 1),
    `% Female` = round(mean(sex == "Female") * 100, 1),
    `Avg Inpatient Admits` = round(mean(inpatient_admits), 2),
    `Avg ED Visits` = round(mean(ed_visits), 2),
    `Avg Total Cost ($)` = round(mean(total_cost), 0),
    `Avg # Chronic Conditions` = round(rowMeans(select(., hypertension, diabetes, copd, depression, cancer)) %>% mean() * 5, 2)
  ) %>%
  mutate(Risk_Group = ifelse(high_risk == 1, "High Risk", "Not High Risk")) %>%
  select(Risk_Group, everything(), -high_risk)

# Render it nicely
risk_profile %>%
  gt() %>%
  tab_header(
    title = "High Risk vs Not High Risk: Key Characteristics"
  ) %>%
  cols_label(
    `Mean Age` = "Mean Age",
    `% Female` = "% Female",
    `Avg Inpatient Admits` = "Avg Inpatient Admits",
    `Avg ED Visits` = "Avg ED Visits",
    `Avg Total Cost ($)` = "Avg Total Cost ($)",
    `Avg # Chronic Conditions` = "Avg Chronic Conditions"
  ) %>%
  fmt_number(columns = everything(), decimals = 1) %>%
  cols_align(align = "center", columns = everything())
```

-   High-risk patients are more likely to be older, with higher ED costs, and higher inpatient utilization.

These findings support the need for a tiered risk segmentation strategy to prioritize care management resources effectively.

# Risk Score Modeling

```{r model-setup, message=FALSE}

# Clean modeling dataset
model_data <- patient_data %>%
  mutate(
    high_risk = factor(high_risk, labels = c("No", "Yes")),
    sex = factor(sex),
    income_bracket = cut(zip_income, breaks = c(-Inf, 40000, 60000, 80000, Inf),
                         labels = c("<40K", "40–60K", "60–80K", "80K+"))
  ) %>%
  select(high_risk, age, sex, income_bracket, hypertension, diabetes, copd,
         depression, cancer, ed_visits, inpatient_admits, med_adherence_score)

# Train/test split
set.seed(42)
train_index <- createDataPartition(model_data$high_risk, p = 0.75, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]
```

```{r log-mod1}
logit_mod1 <-
  glm(high_risk ~., data = train_data, family = binomial)

library(broom)

glm_results <- tidy(logit_mod1) %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    p.value = round(p.value, 3),
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  select(term, estimate, std.error, p.value, significance)

glm_results %>%
  gt() %>%
  tab_header(title = "Logistic Regression Coefficients: High-Risk Prediction") %>%
  cols_label(
    term = "Predictor",
    estimate = "Estimate",
    std.error = "Std. Error",
    p.value = "P-Value",
    significance = " "
  ) %>%
  tab_source_note("Note: *** p < 0.001, ** p < 0.01, * p < 0.05")
```

```{r mod-performance, message=FALSE}
# Predict probabilities on test set
pred_probs <- predict(logit_mod1, newdata = test_data, type = "response")
pred_class <- ifelse(pred_probs > 0.8, "Yes", "No") %>% factor(levels = c("No", "Yes"))

# Confusion matrix
# Fit confusion matrix as usual
cm <- confusionMatrix(pred_class, test_data$high_risk, positive = "Yes")

# Extract the actual confusion matrix counts
cm_matrix <- as.table(cm$table)
cm_df <- as.data.frame(cm_matrix)

# Optional: Rename for readability
names(cm_df) <- c("Actual", "Predicted", "Count")

# Create clean gt table
cm_df %>%
  tidyr::pivot_wider(names_from = Predicted, values_from = Count) %>%
  gt() %>%
  tab_header(title = "Confusion Matrix: Model Predictions vs Actual") %>%
  cols_label(Actual = "Actual Class", No = "Predicted: No", Yes = "Predicted: Yes") %>%
  cols_align("center", everything())

# ROC Curve
roc_obj <- roc(test_data$high_risk, pred_probs)

roc_data <- data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities
)

ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_line(color = "#002d72", size = 1.2) +
  geom_abline(linetype = "dashed", color = "#fdb813") +
  labs(
    title = "ROC Curve: Predicting High-Risk Patients",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal(base_size = 12) +
  coord_fixed()
```

```{r pred-risk-dist}
test_data$predicted_risk <- pred_probs

ggplot(test_data, aes(x = predicted_risk, fill = high_risk)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("#002d72", "#fdb813")) +
  labs(title = "Predicted Risk Distribution", x = "Predicted Probability of High Risk", fill = "Actual")

```

A logistic regression model was trained to predict high-risk patients based on demographics, chronic conditions, and recent utilization patterns.

-   The model achieved an AUC of `r round(auc(roc_obj), 3)`, indicating modest discriminative ability—sufficient for exploratory stratification, ***though not deployment-ready without further tuning***.
-   Features with strongest predictive value included inpatient admits, diabetes, and COPD.
-   The confusion matrix shows balance between sensitivity and specificity at a 0.8 threshold.
-   The predicted risk distribution highlights some moderate separation between high- and low-risk patients, enabling future segmentation.
-   ***Note***: The model predicted all patients as high-risk at the default threshold. This is a common challenge in imbalanced classification and would be addressed through threshold optimization and calibration in a production setting.

This model offers a prototype framework that could inform tiered outreach strategies under value-based care, pending validation with real patient data.

# Cost Savings & ROI Estimation

```{r}
# Target top 10% predicted risk
cutoff <- quantile(test_data$predicted_risk, 0.9)
target_group <- test_data %>%
  filter(predicted_risk >= cutoff)

# Assumptions
avg_savings_pct <- 0.12        # 12% cost savings per targeted patient
intervention_cost <- 400       # per patient per year

# Baseline cost of targeted group
baseline_total_cost <- mean(patient_data$total_cost)  # population-wide
target_cost <- sum(target_group$predicted_risk * baseline_total_cost)

# Estimated savings + cost
estimated_savings <- target_cost * avg_savings_pct
total_program_cost <- nrow(target_group) * intervention_cost
estimated_roi <- (estimated_savings - total_program_cost) / total_program_cost

# Summary output
roi_summary <- data.frame(
  `Patients Targeted` = nrow(target_group),
  `Avg Cost (Est)` = round(target_cost / nrow(target_group), 0),
  `Est. Savings per Patient` = round(avg_savings_pct * target_cost / nrow(target_group), 0),
  `Total Estimated Savings` = round(estimated_savings, 0),
  `Total Program Cost` = round(total_program_cost, 0),
  `ROI (12-Month)` = paste0(round(estimated_roi * 100, 1), "%")
)

roi_summary %>%
  gt() %>%
  tab_header(title = "Estimated ROI: Targeted Care Management Program") %>%
  cols_label(
    Patients.Targeted = "Patients Targeted",
    Avg.Cost..Est. = "Avg Cost (Est.)",
    Est..Savings.per.Patient = "Est. Savings/Patient",
    Total.Estimated.Savings = "Total Estimated Savings",
    Total.Program.Cost = "Program Cost",
    ROI..12.Month. = "ROI (12-Month)"
  ) %>%
  fmt_currency(columns = c(2:5), decimals = 0) %>%
  cols_align("center", everything())
```

To quantify business value, we modeled a care management program targeting the top 10% of predicted high-risk patients.

-   **Patients Targeted:** `r nrow(target_group)`
-   **Estimated 12-Month Cost Savings:** `r scales::dollar(round(estimated_savings, 0))`
-   **Program Cost:** `r scales::dollar(round(total_program_cost, 0))`
-   **Projected ROI:** `r round(estimated_roi * 100, 1)`%

Assuming an average savings of +6.7% in total cost of care, we see some signal in our methods of risk-tiering, consistent with benchmarks for high-impact care coordination efforts.

```{r}
# Sample 5 patients from top 10% predicted risk
sample_patients <- target_group %>%
  select(age, sex, ed_visits, inpatient_admits, 
         med_adherence_score, predicted_risk) %>%
  arrange(desc(predicted_risk)) %>%
  head(5) %>%
  mutate(
    predicted_risk = round(predicted_risk, 2),
    med_adherence_score = round(med_adherence_score, 2)
  )

# Format for presentation
sample_patients %>%
  gt() %>%
  tab_header(title = "Patients Flagged for High-Risk Outreach") %>%
  cols_label(
    age = "Age",
    sex = "Sex",
    ed_visits = "ED Visits",
    inpatient_admits = "Inpatient Admits",
    med_adherence_score = "Med Adherence Score",
    predicted_risk = "Predicted Risk Score"
  ) %>%
  cols_align("center", everything())
```

The table below shows a sample of five patients identified in the top 10% of predicted risk. These individuals share patterns of elevated utilization (e.g., repeat admissions), low-to-moderate medication adherence, and higher overall risk scores.

Targeted care management—such as intensive case reviews or social work referrals—could help reduce unnecessary admissions and improve adherence, supporting both cost containment and clinical outcomes.

This micro-level lens highlights how predictive modeling informs actionable, patient-level decisions under value-based care.

# Strategic Implications

This analysis demonstrates how predictive analytics can enable targeted care management and improve performance under value-based care agreements. By identifying high-risk patients in advance, our team might be able to:

-   Reduce preventable admissions and emergency department visits
-   Allocate care coordination resources more efficiently
-   Quantify and track impact on total cost of care
-   Align operational priorities with population health goals

The risk model provides a scalable foundation for tiered interventions, prospective outreach, and continuous ROI measurement—all aligned with the analytics and leadership responsibilities of the AVP role.

# Conclusion

This project highlights a practical and scalable approach to value-based performance optimization. [**If I were serving this team with access to real patient data, this framework could be piloted and adapted to Ochsner’s specific contracts, care models, and operational realities.**]{.underline}

I look forward to your consideration and talking soon.
